---
title: "Finance with {data.table}"
format: gfm
---

Just a place to store some code snippets and notes on finance with
using the latest `data.table` package. Requires `data.table >= 1.18.0`.

Load the required libraries:

```{r}
library(clock)
library(data.table)
library(ggplot2)
```

## Portfolio Management

#### Generate data

Generate some fake stock prices for a few tickers.

```{r}
set.seed(1994)

generate_prices = function(ticker, start_date, end_date) {
  dates = seq(as.Date(start_date), as.Date(end_date), by = "1 day")
  n = length(dates)
  prices = cumprod(1 + rnorm(n, mean = 0.0005, sd = 0.01)) * 100
  data.table(
    ticker = ticker,
    date = dates,
    price = prices
  )
}

generate_benchmark = function(start_date, end_date) {
  dates = seq(as.Date(start_date), as.Date(end_date), by = "1 day")
  n = length(dates)
  prices = cumprod(1 + rnorm(n, mean = 0.0003, sd = 0.008)) * 3000
  data.table(
    ticker = "SP500",
    date = dates,
    price = prices
  )
}

ticker = c("AAPL", "GOOGL", "MSFT", "AMZN")
start_date = "2015-01-01"
end_date = Sys.Date()

dt = rbindlist(lapply(ticker, generate_prices, start_date, end_date))
alloc = data.table(
  ticker = ticker,
  weight = c(0.4, 0.3, 0.2, 0.1),
  sector = c("Technology", "Technology", "Technology", "Consumer Cyclical"),
  country = c("USA", "USA", "USA", "USA")
)
dt = dt[alloc, on = "ticker"]
head(dt)
```

#### Holdings

```{r}
holdings = dt |>
  _[,
    .(start_price = first(price), current_price = last(price), weight = first(weight)),
    by = ticker
  ] |>
  _[, let(
    value = current_price * weight,
    abs_change = current_price - start_price,
    rel_change = current_price / start_price - 1
  )] |>
  _[, rel_weight := value / sum(value)]
holdings
```

#### Portfolio Composition

```{r}
#| dpi = 300
ggplot(holdings, aes(x = "", y = rel_weight, fill = ticker)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = scales::percent(rel_weight, accuracy = 0.1)),
    position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Composition") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_blank()
  )
```

#### Calculate returns

```{r}
logret = function(x) {
  x = log(x)
  x - shift(x)
}

dt = dt |>
  setorder(ticker, date) |>
  _[, let(ret = price / shift(price) - 1, log_ret = logret(price)), by = ticker] |>
  na.omit("ret") |>
  _[, let(wret = ret * weight, value = price * weight)]
head(dt)
```

```{r}
#| dpi = 300
dt |>
  _[date >= add_months(end_date, -12L), .(value = sum(value)), by = date] |>
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank()
  ) +
  labs(title = "Portfolio Value")
```

#### Calculate weekly, monthly and yearly returns

Return for each instrument:

```{r}
ret_week = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, year(date), week(date))]
ret_month = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, yearmon(date))]
ret_year = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, year(date))]
head(ret_year)
```

Return for the portfolio:

```{r}
port_ret_week = dt[, .(ret = prod(1 + wret) - 1), by = .(year(date), week(date))]
port_ret_month = dt[, .(ret = prod(1 + wret) - 1), by = .(yearmon(date))]
port_ret_year = dt[, .(ret = prod(1 + wret) - 1), by = year(date)]
head(port_ret_year)
```

#### Compare performance with a benchmark

Calculat the benchmark return:

```{r}
bmr = generate_benchmark(start_date, end_date) |>
  setorder(date) |>
  _[, ret := price / shift(price) - 1] |>
  na.omit("ret")

port = dt |>
  _[, .(ret = prod(1 + wret) - 1, ticker = "Portfolio"), by = date] |>
  rbind(bmr[, .(ticker, date, ret)]) |>
  setorder(ticker, date) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, ticker := replace(ticker, ticker != "Portfolio", "Benchmark")]
```

Compare the portfolio with the benchmark performance:

```{r}
#| dpi = 300
port |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date, y = cum_ret, color = ticker)) +
  geom_line() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 2)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark")
```

Or turn it into a wide-format and display the performance as an area chart:

```{r}
#| dpi = 300
perf = port |>
  dcast(date ~ ticker, value.var = "cum_ret") |>
  setnames(tolower) |>
  _[, diff := portfolio - benchmark]

perf |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date)) +
  geom_ribbon(
    aes(
      ymin = pmin(portfolio, benchmark),
      ymax = pmax(portfolio, benchmark),
      fill = diff > 0
    ),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  geom_line(aes(y = portfolio, color = "Portfolio")) +
  geom_line(aes(y = benchmark, color = "Benchmark")) +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_fill_manual(values = c("TRUE" = "#00A651", "FALSE" = "#FF0000")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 2L)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

```{r, eval = FALSE}
perf |>
  _[
    date >= "2022-01-10",
    .(
      benchmark = last(benchmark) - first(benchmark),
      portfolio = last(portfolio) - first(portfolio)
    ),
    by = .(year(date))
  ]
```

#### Analyse the portfolio exposure

```{r}
exposure = dt |>
  _[, .(value = sum(value)), by = .(date, sector)] |>
  _[, weight := value / sum(value), by = date]
head(exposure)
```

Exposure by sector over time:

```{r}
#| dpi = 300
exposure |>
  _[date >= add_months(end_date, -12L)] |>
  ggplot(aes(x = date, y = weight, fill = sector)) +
  geom_area() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Exposure") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

#### Calculate volatility

Note this is scaling volatility by $\sqrt{h}$, which has some shortcomings,
see for example
[Diebold et.al. (1996)](https://www.sas.upenn.edu/~fdiebold/papers/paper18/dsi.pdf&ved=2ahUKEwjM2P-7jfGKAxUkBdsEHcrTCAkQFnoECBcQAQ&usg=AOvVaw36skVdLjP1SwTgB6J1rdnz).


```{r}
vola = dt |>
  _[, .(daily_vola = sd(log_ret)), by = .(ticker, year(date))] |>
  _[, let(
    weekly_vola = daily_vola * sqrt(5),
    monthly_vola = daily_vola * sqrt(21),
    yearly_vola = daily_vola * sqrt(252)
  )]
head(vola)
```

#### Rolling volatility

```{r}
#| dpi = 300
window = 63L # ~3 months
port_daily = dt[, .(ret = sum(wret)), by = date]
port_daily[, roll_vola := frollsd(ret, window) * sqrt(252)]

port_daily |>
  na.omit("roll_vola") |>
  ggplot(aes(x = date, y = roll_vola)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Rolling Annualized Volatility (63-day)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank()
  )
```

#### Sharpe ratio

The Sharpe ratio measures risk-adjusted return:

$$
S = \frac{R_p - R_f}{\sigma_p}
$$

```{r}
rf = 0.04 / 252 # daily risk-free rate (assuming 4% annual)
port_daily = dt[, .(ret = sum(wret)), by = date]
sharpe = port_daily[, (mean(ret) - rf) / sd(ret) * sqrt(252)]
sharpe
```

#### Sortino ratio

The Sortino ratio replaces total volatility with downside deviation:

$$
So = \frac{R_p - R_f}{\sigma_d}
$$

```{r}
sortino = port_daily[, (mean(ret) - rf) / sd(pmin(ret - rf, 0)) * sqrt(252)]
sortino
```

#### Value at Risk

Historical VaR at the 95% and 99% confidence levels:

```{r}
port_daily[, .(VaR_95 = quantile(ret, 0.05), VaR_99 = quantile(ret, 0.01))]
```

#### Portfolio risk

Portfolio risk is defined as:

$$
\sigma_p = \sqrt{w^T \Sigma w}
$$

```{r}
wgt = alloc$weight
cov_mat = dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cov(use = "pairwise.complete.obs")
port_risk = as.numeric(sqrt(t(wgt) %*% cov_mat %*% wgt))
port_risk
```

#### Drawdown

Maximum Drawdown is defined as follows:

$$
MDD = \max_{i \leq j} \left( \frac{V_j - V_i}{V_i} \right)
$$

Instrument drawdown:

```{r}
drawdown = copy(dt) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, drawdown := (cum_ret - cummax(cum_ret)), by = ticker]
head(drawdown)
```

Portfolio drawdown:

```{r}
drawdown = dt |>
  _[, .(wret = sum(wret)), by = date] |>
  _[, cum_ret := cumprod(1 + wret) - 1] |>
  _[, drawdown := (cum_ret - cummax(cum_ret))]
head(drawdown)
```

```{r}
drawdown[drawdown < 0, .(min_drawdown = min(drawdown), avg_drawdown = mean(drawdown))]
```

#### Tracking error

Tracking error measures how closely a portfolio follows its benchmark:

$$
TE = \sqrt{\frac{1}{N-1} \sum_{i=1}^{N} (r_{p,i} - r_{b,i})^2}
$$

```{r}
te = dt |>
  _[, .(port_ret = sum(wret)), by = date] |>
  _[bmr[, .(date, bmr_ret = ret)], on = "date", nomatch = NULL] |>
  _[, diff := port_ret - bmr_ret]

te[, .(
  daily_te = sd(diff),
  annual_te = sd(diff) * sqrt(252)
)]
```

#### Beta and Alpha

Beta measures the portfolio's sensitivity to the benchmark, Alpha the excess return:

$$
r_p = \alpha + \beta \cdot r_b + \epsilon
$$

```{r}
fit = te[, lm(port_ret ~ bmr_ret)]
coefs = coef(fit)
data.table(alpha = coefs[1L] * 252, beta = coefs[2L])
```

#### Correlation matrix

```{r}
dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cor(use = "pairwise.complete.obs") |>
  round(3)
```
