---
title: "Finance with {data.table}"
format:
  gfm:
    toc: true
    toc-depth: 4
---

Just a place to store some code snippets and notes on finance with
using the latest `data.table` package. Requires `data.table >= 1.18.0`.

Load the required libraries:

```{r}
#| message: false
library(clock)
library(data.table)
library(ggplot2)

theme_finance = theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    legend.title = element_blank()
  )
```

## Portfolio Management

#### Generate data

We need a reproducible universe to work with, so rather than download real
quotes we simulate them from a known return-generating process. Each instrument
$i$ earns a daily return driven by a common market factor, a sector factor
shared by its peers, and idiosyncratic noise:

$$
r_{i,t} = \alpha_i + \beta_i \, r_{m,t} + r_{s(i),t} + \epsilon_{i,t},
\qquad \epsilon_{i,t} \sim \mathcal{N}(0, \sigma_i^2)
$$

Prices are then the cumulative product of these returns starting from a base
level $P_{i,0}$:

$$
P_{i,t} = P_{i,0} \prod_{u=1}^{t} (1 + r_{i,u})
$$

Because we choose $\alpha_i$, $\beta_i$ and the factors ourselves, we can later
check that the estimators below recover them.

```{r}
set.seed(1994)

start_date = "2015-01-01"
end_date = Sys.Date()
dates = seq(as.Date(start_date), end_date, by = "1 day")
n = length(dates)

market_ret = rnorm(n, mean = 0.0003, sd = 0.008)
sector_ret = list(
  technology = rnorm(n, mean = 0, sd = 0.004),
  consumer_cyclical = rnorm(n, mean = 0, sd = 0.005)
)

alloc = data.table(
  ticker = c("AAPL", "GOOGL", "MSFT", "AMZN"),
  weight = c(0.4, 0.3, 0.2, 0.1),
  sector = c("technology", "technology", "technology", "consumer_cyclical"),
  country = c("USA", "USA", "USA", "USA"),
  alpha = c(0.0004, 0.0002, 0.0003, 0.0001),
  beta = c(1.1, 1.3, 0.9, 1.4),
  idio_sd = c(0.010, 0.012, 0.009, 0.014)
)

generate_prices = function(alpha, beta, sector, idio_sd) {
  ret = alpha + beta * market_ret + sector_ret[[sector]] + rnorm(n, sd = idio_sd)
  data.table(date = dates, price = cumprod(1 + ret) * 100)
}

generate_benchmark = function() {
  data.table(ticker = "SP500", date = dates, price = cumprod(1 + market_ret) * 3000)
}

dt = alloc |>
  _[, generate_prices(alpha, beta, sector, idio_sd), by = ticker] |>
  _[alloc[, .(ticker, weight, sector, country)], on = "ticker"]
head(dt)
```

#### Holdings

A holdings table is the snapshot view of the portfolio: where does the money sit
right now? We collapse the price history to the first and last observation per
instrument, value each position at its current price, and express it as a share
of the total. The position value and its portfolio weight are

$$
v_i = w_i \, P_{i,T}, \qquad \tilde{w}_i = \frac{v_i}{\sum_j v_j}
$$

where $w_i$ is the target weight and $\tilde{w}_i$ the realized weight after
prices have drifted apart.

```{r}
holdings = dt |>
  _[,
    .(start_price = first(price), current_price = last(price), weight = first(weight)),
    by = ticker
  ] |>
  _[, let(
    value = current_price * weight,
    abs_change = current_price - start_price,
    rel_change = current_price / start_price - 1
  )] |>
  _[, rel_weight := value / sum(value)]
head(holdings)
```

#### Portfolio Composition

To see concentration at a glance we plot the realized weights $\tilde{w}_i$ as a
pie chart. Because $\sum_i \tilde{w}_i = 1$, each slice is the fraction of total
portfolio value held in that instrument.

```{r}
#| dpi: 300
ggplot(holdings, aes(x = "", y = rel_weight, fill = ticker)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = scales::percent(rel_weight, accuracy = 0.1)),
    position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Composition") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_blank()
  )
```

#### Calculate returns

Returns, not prices, are the unit of analysis for almost everything that
follows, because they are roughly stationary and comparable across instruments.
We use two definitions — the simple (arithmetic) return and the log return:

$$
r_t = \frac{P_t}{P_{t-1}} - 1, \qquad \ell_t = \ln \frac{P_t}{P_{t-1}} = \ln(1 + r_t)
$$

Simple returns aggregate cleanly across instruments (the portfolio return is a
weighted sum), while log returns aggregate cleanly across time (they add up over
periods). The position's contribution to the portfolio is the weighted return
$w_i \, r_{i,t}$.

```{r}
logret = function(x) {
  x = log(x)
  x - shift(x)
}

dt = dt |>
  setorder(ticker, date) |>
  _[, let(ret = price / shift(price) - 1, log_ret = logret(price)), by = ticker] |>
  na.omit("ret") |>
  _[, let(wret = ret * weight, value = price * weight)]
head(dt)
```

```{r}
#| dpi: 300
dt |>
  _[date >= add_months(end_date, -12L), .(value = sum(value)), by = date] |>
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  theme_finance +
  labs(title = "Portfolio Value")
```

#### Calculate weekly, monthly and yearly returns

Daily returns compound into longer-horizon returns geometrically, not by
addition: holding through a window $[t_1, t_2]$ multiplies the per-period growth
factors,

$$
r_{[t_1, t_2]} = \prod_{t=t_1}^{t_2} (1 + r_t) - 1
$$

The same compounding applies to the portfolio, whose daily return is the
weighted sum of its constituents, $r_{p,t} = \sum_i w_i \, r_{i,t}$.

Return for each instrument:

```{r}
ret_week = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, isoyear(date), isoweek(date))]
ret_month = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, yearmon(date))]
ret_year = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, year(date))]
head(ret_year)
```

Return for the portfolio:

```{r}
port_daily = dt[, .(ret = sum(wret)), by = date]
port_ret_week = port_daily[, .(ret = prod(1 + ret) - 1), by = .(isoyear(date), isoweek(date))]
port_ret_month = port_daily[, .(ret = prod(1 + ret) - 1), by = .(yearmon(date))]
port_ret_year = port_daily[, .(ret = prod(1 + ret) - 1), by = year(date)]
head(port_ret_year)
```

#### Monthly return heatmap

Compounding the daily portfolio return within each calendar month gives one
number per month, $r_{p,[m]} = \prod_{t \in m} (1 + r_{p,t}) - 1$. Laying these
out as a year × month grid makes seasonality and the worst months jump out at a
glance.

```{r}
#| dpi: 300
port_ret_month_dt = port_daily[,
  .(ret = prod(1 + ret) - 1),
  by = .(year = year(date), month = month(date))
]

ggplot(
  port_ret_month_dt,
  aes(x = fctr(month), y = fctr(year, rev = TRUE), fill = ret)
) +
  geom_tile() +
  geom_text(aes(label = scales::percent(ret, accuracy = 0.1)), size = 2.5) +
  scale_fill_gradient2(
    low = "#FF0000",
    mid = "white",
    high = "darkblue",
    midpoint = 0,
    labels = scales::label_percent()
  ) +
  scale_x_discrete(labels = month.abb) +
  labs(title = "Monthly Portfolio Returns") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_blank()
  )
```

#### Compare performance with a benchmark

A return number in isolation says little; what matters is performance *relative*
to an investable alternative such as the market index. We track the cumulative
(wealth) return of both, which is what an investor's account would actually show:

$$
R_t = \prod_{u=1}^{t} (1 + r_u) - 1
$$

Calculate the benchmark return:

```{r}
bmr = generate_benchmark() |>
  setorder(date) |>
  _[, ret := price / shift(price) - 1] |>
  na.omit("ret")

port = dt |>
  _[, .(ret = sum(wret), ticker = "Portfolio"), by = date] |>
  rbind(bmr[, .(ticker, date, ret)]) |>
  setorder(ticker, date) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, ticker := fifelse(ticker == "Portfolio", ticker, "Benchmark")]
```

Compare the portfolio with the benchmark performance:

```{r}
#| dpi: 300
port |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date, y = cum_ret, color = ticker)) +
  geom_line() +
  theme_finance +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark")
```

Or turn it into a wide-format and display the performance as an area chart:

```{r}
#| dpi: 300
perf = port |>
  dcast(date ~ ticker, value.var = "cum_ret") |>
  setnames(tolower) |>
  _[, diff := portfolio - benchmark]

perf |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date)) +
  geom_ribbon(
    aes(
      ymin = pmin(portfolio, benchmark),
      ymax = pmax(portfolio, benchmark),
      fill = diff > 0
    ),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  geom_line(aes(y = portfolio, color = "Portfolio")) +
  geom_line(aes(y = benchmark, color = "Benchmark")) +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_fill_manual(values = c("TRUE" = "#00A651", "FALSE" = "#FF0000")) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1L)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Analyse the portfolio exposure

Risk is rarely spread evenly across sectors, so we decompose the portfolio value
by sector at each date. The exposure of sector $s$ is the fraction of total value
held in its members, and these fractions drift over time as prices move:

$$
w_{s,t} = \frac{\sum_{i \in s} v_{i,t}}{\sum_j v_{j,t}}
$$

```{r}
exposure = dt |>
  _[, .(value = sum(value)), by = .(date, sector)] |>
  _[, weight := value / sum(value), by = date]
head(exposure)
```

Exposure by sector over time:

```{r}
#| dpi: 300
exposure |>
  _[date >= add_months(end_date, -12L)] |>
  ggplot(aes(x = date, y = weight, fill = sector)) +
  geom_area() +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Exposure") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Calculate volatility

Volatility quantifies how dispersed returns are around their mean — the standard
measure of risk. We estimate it as the sample standard deviation of (log)
returns over a period:

$$
\sigma_d = \sqrt{\frac{1}{N - 1} \sum_{t=1}^{N} (\ell_t - \bar{\ell})^2}
$$

To compare horizons we annualize. Under the (strong) assumption of i.i.d.
returns, variance grows linearly with time, so volatility grows with the square
root of the number of periods $h$:

$$
\sigma_h = \sigma_d \sqrt{h}
$$

Note this $\sqrt{h}$ scaling has some shortcomings, see for example
[Diebold et.al. (1996)](https://www.sas.upenn.edu/~fdiebold/papers/paper18/dsi.pdf).


```{r}
vola = dt |>
  _[, .(daily_vola = sd(log_ret)), by = .(ticker, year(date))] |>
  _[, let(
    weekly_vola = daily_vola * sqrt(5),
    monthly_vola = daily_vola * sqrt(21),
    yearly_vola = daily_vola * sqrt(252)
  )]
head(vola)
```

#### Rolling volatility

A single full-sample number hides how risk changes over time. Recomputing the
annualized standard deviation over a trailing window of $w$ days turns
volatility into a time series that reveals calm and turbulent regimes:

$$
\sigma_t^{(w)} = \sqrt{252} \cdot \sqrt{\frac{1}{w - 1} \sum_{u=t-w+1}^{t} (r_u - \bar{r}_t)^2}
$$

```{r}
#| dpi: 300
window = 63L # ~3 months
port_daily[, roll_vola := frollsd(ret, window) * sqrt(252)]

port_daily |>
  na.omit("roll_vola") |>
  ggplot(aes(x = date, y = roll_vola)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = "Rolling Annualized Volatility (63-day)") +
  theme_finance
```

#### Sharpe ratio

The Sharpe ratio measures risk-adjusted return:

$$
S = \frac{R_p - R_f}{\sigma_p}
$$

```{r}
rf = 0.04 / 252 # daily risk-free rate (assuming 4% annual)
sharpe = port_daily[, (mean(ret) - rf) / sd(ret) * sqrt(252)]
sharpe
```

#### Sortino ratio

Investors do not mind upside volatility — only losses. The Sortino ratio
therefore replaces total volatility with the *downside deviation*, which
penalizes only returns below the target (here the risk-free rate):

$$
So = \frac{R_p - R_f}{\sigma_d},
\qquad \sigma_d = \sqrt{\frac{1}{N} \sum_{t=1}^{N} \min(r_t - R_f, \, 0)^2}
$$

```{r}
sortino = port_daily[, (mean(ret) - rf) / sqrt(mean(pmin(ret - rf, 0)^2)) * sqrt(252)]
sortino
```

#### Rolling Sharpe

Just like volatility, risk-adjusted return is not constant. Evaluating the
Sharpe ratio over a trailing window shows when the portfolio was actually being
paid for the risk it took:

$$
S_t^{(w)} = \sqrt{252} \cdot \frac{\bar{r}_t - R_f}{\sigma_t^{(w)} / \sqrt{252}}
$$

```{r}
#| dpi: 300
port_daily[, roll_sharpe := (frollmean(ret, window) - rf) / frollsd(ret, window) * sqrt(252)]

port_daily |>
  na.omit("roll_sharpe") |>
  ggplot(aes(x = date, y = roll_sharpe)) +
  geom_line() +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(title = "Rolling Sharpe Ratio (63-day)") +
  theme_finance
```

#### Value at Risk

Value at Risk answers a regulator's question: "over one day, how bad can it
plausibly get?" At confidence level $1 - \alpha$ it is the loss that returns will
not exceed with probability $1 - \alpha$, i.e. the lower $\alpha$-quantile of the
return distribution:

$$
\mathrm{VaR}_{1-\alpha} = Q_r(\alpha), \qquad P(r \le Q_r(\alpha)) = \alpha
$$

The *historical* estimate makes no distributional assumption — it simply reads
the empirical quantile off the realized returns.

Historical VaR at the 95% and 99% confidence levels:

```{r}
port_daily[, .(VaR_95 = quantile(ret, 0.05), VaR_99 = quantile(ret, 0.01))]
```

#### Expected Shortfall (CVaR)

VaR tells you a threshold but is silent about *how bad* the tail beyond it is,
and it is not sub-additive (diversification can paradoxically increase it).
Expected Shortfall fixes both by averaging the losses that fall beyond the VaR
quantile:

$$
\mathrm{ES}_{1-\alpha} = \mathbb{E}\!\left[\, r \mid r \le Q_r(\alpha) \,\right]
$$

Average loss beyond VaR:

```{r}
port_daily[, .(
  CVaR_95 = mean(ret[ret <= quantile(ret, 0.05)]),
  CVaR_99 = mean(ret[ret <= quantile(ret, 0.01)])
)]
```

#### Portfolio risk

Portfolio volatility is *not* the weighted average of the individual
volatilities — correlations matter, which is the whole point of
diversification. With weight vector $w$ and return covariance matrix $\Sigma$,
risk is the quadratic form

$$
\sigma_p = \sqrt{w^\top \Sigma w}
$$

The off-diagonal terms of $\Sigma$ are what let imperfectly correlated assets
partially cancel each other's risk.

```{r}
cov_mat = dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cov(use = "pairwise.complete.obs")
wgt = alloc[colnames(cov_mat), weight, on = "ticker"]
port_risk = as.numeric(sqrt(t(wgt) %*% cov_mat %*% wgt))
port_risk
```

#### Drawdown

Volatility treats gains and losses symmetrically, but what actually makes
investors capitulate is the peak-to-trough decline. The drawdown at time $t$
measures how far the cumulative value $V_t$ has fallen below its running
all-time high, and the maximum drawdown is the worst such drop over the whole
path:

$$
D_t = \frac{V_t}{\max_{u \le t} V_u} - 1, \qquad
MDD = \max_{i \leq j} \left( \frac{V_i - V_j}{V_i} \right)
$$

Instrument drawdown:

```{r}
drawdown = copy(dt) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, drawdown := (1 + cum_ret) / cummax(1 + cum_ret) - 1, by = ticker]
head(drawdown)
```

Portfolio drawdown:

```{r}
drawdown = dt |>
  _[, .(wret = sum(wret)), by = date] |>
  _[, cum_ret := cumprod(1 + wret) - 1] |>
  _[, drawdown := (1 + cum_ret) / cummax(1 + cum_ret) - 1]
head(drawdown)
```

```{r}
drawdown[drawdown < 0, .(min_drawdown = min(drawdown), avg_drawdown = mean(drawdown))]
```

#### Calmar ratio

The Calmar ratio is a risk-adjusted return that uses the maximum drawdown as its
risk measure instead of volatility — appealing because drawdown is the risk
investors actually feel. It is the annualized return earned per unit of worst-case
loss:

$$
\mathrm{Calmar} = \frac{R_{\text{ann}}}{|MDD|}
$$

```{r}
calmar = port_daily[, (mean(ret) * 252) / abs(drawdown[, min(drawdown)])]
calmar
```

#### Tracking error

Tracking error measures how closely a portfolio follows its benchmark:

$$
TE = \sqrt{\frac{1}{N-1} \sum_{i=1}^{N} (r_{p,i} - r_{b,i})^2}
$$

```{r}
te = dt |>
  _[, .(port_ret = sum(wret)), by = date] |>
  _[bmr[, .(date, bmr_ret = ret)], on = "date", nomatch = NULL] |>
  _[, diff := port_ret - bmr_ret]

te[, .(
  daily_te = sd(diff),
  annual_te = sd(diff) * sqrt(252)
)]
```

#### Information ratio

Tracking error alone does not say whether the active bets paid off. The
information ratio is the active-management analogue of the Sharpe ratio: it
divides the *mean* excess return over the benchmark by the tracking error,
measuring skill per unit of active risk taken:

$$
IR = \frac{\overline{r_p - r_b}}{TE}
$$

```{r}
te[, mean(diff) / sd(diff) * sqrt(252)]
```

#### Beta and Alpha

Beta measures the portfolio's sensitivity to the benchmark, Alpha the excess return:

$$
r_p = \alpha + \beta \cdot r_b + \epsilon
$$

```{r}
fit = te[, lm(port_ret ~ bmr_ret)]
coefs = coef(fit)
data.table(alpha = coefs[1L] * 252, beta = coefs[2L])
```

#### Multi-factor model

The single-factor model above explains returns with the benchmark alone, but our
data-generating process used two systematic factors. Adding the sector factor
gives a multi-factor regression whose loadings should recover the parameters we
simulated with:

$$
r_{i,t} = \alpha_i + \beta_i^{m} \, r_{m,t} + \beta_i^{s} \, r_{s(i),t} + \epsilon_{i,t}
$$

Each instrument is driven by the common market factor plus its sector factor, so
we regress every ticker on both and recover its factor loadings. Assemble the
factors and reshape them to long form, then match each ticker to its own sector
factor with a join on `date` and `sector`:

```{r}
factors = data.table(
  date = dates,
  mkt = market_ret,
  technology = sector_ret$technology,
  consumer_cyclical = sector_ret$consumer_cyclical
) |>
  melt(id.vars = c("date", "mkt"), variable.name = "sector", value.name = "sec_ret") |>
  _[, sector := as.character(sector)]

reg = factors[dt, on = .(date, sector), nomatch = NULL]
```

Run one regression per ticker and compare the recovered market beta with the value that
was used to generate the data:

```{r}
loadings = reg[, as.list(coef(lm(ret ~ mkt + sec_ret))), by = ticker] |>
  setnames(c("ticker", "alpha", "mkt_beta", "sector_beta")) |>
  _[, alpha := alpha * 252]
loadings[alloc, on = "ticker", true_beta := i.beta]
loadings
```

#### Rolling market beta

Beta is not static — a company's sensitivity to the market shifts as its business
and leverage change. Estimating it over a trailing window exposes this drift.
Over the window, beta is the ratio of the covariance with the market to the
market variance, exactly the slope of the single-factor regression:

$$
\beta_t = \frac{\mathrm{Cov}_t(r_i, \, r_m)}{\mathrm{Var}_t(r_m)}
$$



```{r}
#| dpi: 300
reg |>
  setorder(ticker, date) |>
  _[,
    let(
      cov_rm = frollmean(ret * mkt, window) - frollmean(ret, window) * frollmean(mkt, window),
      var_m = frollmean(mkt^2, window) - frollmean(mkt, window)^2
    ),
    by = ticker
  ] |>
  _[, roll_beta := cov_rm / var_m] |>
  na.omit("roll_beta") |>
  ggplot(aes(x = date, y = roll_beta, color = ticker)) +
  geom_line() +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Rolling Market Beta (63-day)") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Correlation matrix

The covariance matrix drives portfolio risk, but its entries are hard to read
because they mix scale with co-movement. Normalizing covariance by the two
volatilities strips out scale and leaves a unit-free measure of co-movement in
$[-1, 1]$:

$$
\rho_{ij} = \frac{\Sigma_{ij}}{\sigma_i \, \sigma_j}
$$

Strongly correlated names offer little diversification — a fact the optimization
section exploits directly.

```{r}
cor_mat = dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cor(use = "pairwise.complete.obs")
round(cor_mat, 3)
```

```{r}
#| dpi: 300
cor_dt = as.data.table(cor_mat, keep.rownames = "ticker1") |>
  melt(id.vars = "ticker1", variable.name = "ticker2", value.name = "cor")

ggplot(cor_dt, aes(x = ticker1, y = ticker2, fill = cor)) +
  geom_tile() +
  geom_text(aes(label = round(cor, 2)), size = 3) +
  scale_fill_gradient2(low = "#FF0000", mid = "white", high = "darkblue", midpoint = 0) +
  labs(title = "Correlation Matrix") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```

## Portfolio optimization

The covariance matrix from the *Portfolio risk* section, together with the
expected return of each instrument, is everything we need to find optimal
weights. The closed-form solutions below are *unconstrained*, so weights may turn
negative (short positions); adding a long-only constraint would require a
quadratic-programming solver.

#### Expected returns and annualized covariance

Use the mean `log_ret` per ticker for the expected returns, consistent with the
log-return covariance, and annualize both. Align the return vector to the columns
of `cov_mat` with the same join idiom used earlier for the weights:

```{r}
sigma = cov_mat * 252
mu = dt[, .(mu = mean(log_ret) * 252), by = ticker]
mu_vec = mu[colnames(sigma), mu, on = "ticker"]

prec = solve(sigma) # inverse covariance (precision) matrix
ones = rep(1, nrow(sigma))

port_stats = function(w) {
  list(ret = as.numeric(t(w) %*% mu_vec), vol = as.numeric(sqrt(t(w) %*% sigma %*% w)))
}
```

#### Minimum-variance portfolio

The portfolio with the lowest possible variance has a closed-form solution:

$$
w_{mv} = \frac{\Sigma^{-1} \mathbf{1}}{\mathbf{1}^{\top} \Sigma^{-1} \mathbf{1}}
$$

```{r}
w_mv = prec %*% ones / as.numeric(t(ones) %*% prec %*% ones)
mv = port_stats(w_mv)
mv
```

#### Maximum-Sharpe (tangency) portfolio

The tangency portfolio maximises the Sharpe ratio and points where the capital
market line touches the efficient frontier:

$$
w_{tan} \propto \Sigma^{-1} (\mu - r_f \mathbf{1})
$$

```{r}
rf_ann = rf * 252
w_tan = prec %*% (mu_vec - rf_ann)
w_tan = w_tan / sum(w_tan)
tan = port_stats(w_tan)
c(tan, sharpe = (tan$ret - rf_ann) / tan$vol)
```

#### Compare the weights

```{r}
weights = data.table(
  ticker = colnames(sigma),
  current = alloc[colnames(sigma), weight, on = "ticker"],
  min_var = as.numeric(w_mv),
  tangency = as.numeric(w_tan)
)
weights
```

#### Efficient frontier

Every frontier portfolio can be traced in closed form from three scalars derived
from $\Sigma^{-1}$, $\mu$ and $\mathbf{1}$:

$$
\sigma_p^2(\mu_p) = \frac{A \mu_p^2 - 2 B \mu_p + C}{A C - B^2},
\quad A = \mathbf{1}^{\top} \Sigma^{-1} \mathbf{1},
\quad B = \mathbf{1}^{\top} \Sigma^{-1} \mu,
\quad C = \mu^{\top} \Sigma^{-1} \mu
$$

```{r}
A = as.numeric(t(ones) %*% prec %*% ones)
B = as.numeric(t(ones) %*% prec %*% mu_vec)
C = as.numeric(t(mu_vec) %*% prec %*% mu_vec)
D = A * C - B^2

frontier = data.table(ret = seq(min(mu_vec), tan$ret, length.out = 100L)) |>
  _[, vol := sqrt((A * ret^2 - 2 * B * ret + C) / D)]
head(frontier)
```

Plot the frontier together with the individual instruments and the two optimal
portfolios:

```{r}
#| dpi: 300
assets = data.table(ticker = colnames(sigma), ret = mu_vec, vol = sqrt(diag(sigma)))
specials = data.table(
  label = c("Min variance", "Tangency"),
  ret = c(mv$ret, tan$ret),
  vol = c(mv$vol, tan$vol)
)

ggplot(frontier, aes(x = vol, y = ret)) +
  geom_path(color = "darkblue") +
  geom_point(data = assets, color = "black") +
  geom_text(data = assets, aes(label = ticker), vjust = -1, size = 3) +
  geom_point(data = specials, aes(color = label), size = 3) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_color_manual(values = c("Min variance" = "#00A651", "Tangency" = "#FF0000")) +
  labs(title = "Efficient Frontier", x = "Volatility", y = "Expected return") +
  theme_finance +
  theme(axis.title = element_text(), legend.position = "bottom")
```
