---
title: "Finance with {data.table}"
format: gfm
---

Just a place to store some code snippets and notes on finance with
using the latest `data.table` package. Requires `data.table >= 1.18.0`.

Load the required libraries:

```{r}
#| message: false
library(clock)
library(data.table)
library(ggplot2)

theme_finance = theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.2),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    axis.title = element_blank(),
    legend.title = element_blank()
  )
```

## Portfolio Management

#### Generate data

Generate some fake stock prices for a few tickers.

```{r}
set.seed(1994)

start_date = "2015-01-01"
end_date = Sys.Date()
dates = seq(as.Date(start_date), end_date, by = "1 day")
n = length(dates)

market_ret = rnorm(n, mean = 0.0003, sd = 0.008)
sector_ret = list(
  technology = rnorm(n, mean = 0, sd = 0.004),
  consumer_cyclical = rnorm(n, mean = 0, sd = 0.005)
)

alloc = data.table(
  ticker = c("AAPL", "GOOGL", "MSFT", "AMZN"),
  weight = c(0.4, 0.3, 0.2, 0.1),
  sector = c("technology", "technology", "technology", "consumer_cyclical"),
  country = c("USA", "USA", "USA", "USA"),
  alpha = c(0.0004, 0.0002, 0.0003, 0.0001),
  beta = c(1.1, 1.3, 0.9, 1.4),
  idio_sd = c(0.010, 0.012, 0.009, 0.014)
)

generate_prices = function(alpha, beta, sector, idio_sd) {
  ret = alpha + beta * market_ret + sector_ret[[sector]] + rnorm(n, sd = idio_sd)
  data.table(date = dates, price = cumprod(1 + ret) * 100)
}

generate_benchmark = function() {
  data.table(ticker = "SP500", date = dates, price = cumprod(1 + market_ret) * 3000)
}

dt = alloc |>
  _[, generate_prices(alpha, beta, sector, idio_sd), by = ticker] |>
  _[alloc[, .(ticker, weight, sector, country)], on = "ticker"]
head(dt)
```

#### Holdings

```{r}
holdings = dt |>
  _[,
    .(start_price = first(price), current_price = last(price), weight = first(weight)),
    by = ticker
  ] |>
  _[, let(
    value = current_price * weight,
    abs_change = current_price - start_price,
    rel_change = current_price / start_price - 1
  )] |>
  _[, rel_weight := value / sum(value)]
head(holdings)
```

#### Portfolio Composition

```{r}
#| dpi: 300
ggplot(holdings, aes(x = "", y = rel_weight, fill = ticker)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = scales::percent(rel_weight, accuracy = 0.1)),
    position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Composition") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_blank()
  )
```

#### Calculate returns

```{r}
logret = function(x) {
  x = log(x)
  x - shift(x)
}

dt = dt |>
  setorder(ticker, date) |>
  _[, let(ret = price / shift(price) - 1, log_ret = logret(price)), by = ticker] |>
  na.omit("ret") |>
  _[, let(wret = ret * weight, value = price * weight)]
head(dt)
```

```{r}
#| dpi: 300
dt |>
  _[date >= add_months(end_date, -12L), .(value = sum(value)), by = date] |>
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  theme_finance +
  labs(title = "Portfolio Value")
```

#### Calculate weekly, monthly and yearly returns

Return for each instrument:

```{r}
ret_week = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, isoyear(date), isoweek(date))]
ret_month = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, yearmon(date))]
ret_year = dt[, .(ret = prod(1 + ret) - 1), by = .(ticker, year(date))]
head(ret_year)
```

Return for the portfolio:

```{r}
port_daily = dt[, .(ret = sum(wret)), by = date]
port_ret_week = port_daily[, .(ret = prod(1 + ret) - 1), by = .(isoyear(date), isoweek(date))]
port_ret_month = port_daily[, .(ret = prod(1 + ret) - 1), by = .(yearmon(date))]
port_ret_year = port_daily[, .(ret = prod(1 + ret) - 1), by = year(date)]
head(port_ret_year)
```

#### Monthly return heatmap

```{r}
#| dpi: 300
port_ret_month_dt = port_daily[,
  .(ret = prod(1 + ret) - 1),
  by = .(year = year(date), month = month(date))
]

ggplot(
  port_ret_month_dt,
  aes(x = fctr(month), y = fctr(year, rev = TRUE), fill = ret)
) +
  geom_tile() +
  geom_text(aes(label = scales::percent(ret, accuracy = 0.1)), size = 2.5) +
  scale_fill_gradient2(
    low = "#FF0000",
    mid = "white",
    high = "darkblue",
    midpoint = 0,
    labels = scales::label_percent()
  ) +
  scale_x_discrete(labels = month.abb) +
  labs(title = "Monthly Portfolio Returns") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_blank()
  )
```

#### Compare performance with a benchmark

Calculate the benchmark return:

```{r}
bmr = generate_benchmark() |>
  setorder(date) |>
  _[, ret := price / shift(price) - 1] |>
  na.omit("ret")

port = dt |>
  _[, .(ret = sum(wret), ticker = "Portfolio"), by = date] |>
  rbind(bmr[, .(ticker, date, ret)]) |>
  setorder(ticker, date) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, ticker := fifelse(ticker == "Portfolio", ticker, "Benchmark")]
```

Compare the portfolio with the benchmark performance:

```{r}
#| dpi: 300
port |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date, y = cum_ret, color = ticker)) +
  geom_line() +
  theme_finance +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark")
```

Or turn it into a wide-format and display the performance as an area chart:

```{r}
#| dpi: 300
perf = port |>
  dcast(date ~ ticker, value.var = "cum_ret") |>
  setnames(tolower) |>
  _[, diff := portfolio - benchmark]

perf |>
  _[date > "2021-01-01"] |>
  ggplot(aes(x = date)) +
  geom_ribbon(
    aes(
      ymin = pmin(portfolio, benchmark),
      ymax = pmax(portfolio, benchmark),
      fill = diff > 0
    ),
    alpha = 0.4,
    show.legend = FALSE
  ) +
  geom_line(aes(y = portfolio, color = "Portfolio")) +
  geom_line(aes(y = benchmark, color = "Benchmark")) +
  scale_color_manual(values = c("Portfolio" = "darkblue", "Benchmark" = "black")) +
  scale_fill_manual(values = c("TRUE" = "#00A651", "FALSE" = "#FF0000")) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1L)) +
  labs(title = "Cumulative Return: Portfolio vs. Benchmark") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Analyse the portfolio exposure

```{r}
exposure = dt |>
  _[, .(value = sum(value)), by = .(date, sector)] |>
  _[, weight := value / sum(value), by = date]
head(exposure)
```

Exposure by sector over time:

```{r}
#| dpi: 300
exposure |>
  _[date >= add_months(end_date, -12L)] |>
  ggplot(aes(x = date, y = weight, fill = sector)) +
  geom_area() +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Portfolio Exposure") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Calculate volatility

Note this is scaling volatility by $\sqrt{h}$, which has some shortcomings,
see for example
[Diebold et.al. (1996)](https://www.sas.upenn.edu/~fdiebold/papers/paper18/dsi.pdf).


```{r}
vola = dt |>
  _[, .(daily_vola = sd(log_ret)), by = .(ticker, year(date))] |>
  _[, let(
    weekly_vola = daily_vola * sqrt(5),
    monthly_vola = daily_vola * sqrt(21),
    yearly_vola = daily_vola * sqrt(252)
  )]
head(vola)
```

#### Rolling volatility

```{r}
#| dpi: 300
window = 63L # ~3 months
port_daily[, roll_vola := frollsd(ret, window) * sqrt(252)]

port_daily |>
  na.omit("roll_vola") |>
  ggplot(aes(x = date, y = roll_vola)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(title = "Rolling Annualized Volatility (63-day)") +
  theme_finance
```

#### Sharpe ratio

The Sharpe ratio measures risk-adjusted return:

$$
S = \frac{R_p - R_f}{\sigma_p}
$$

```{r}
rf = 0.04 / 252 # daily risk-free rate (assuming 4% annual)
sharpe = port_daily[, (mean(ret) - rf) / sd(ret) * sqrt(252)]
sharpe
```

#### Sortino ratio

The Sortino ratio replaces total volatility with downside deviation:

$$
So = \frac{R_p - R_f}{\sigma_d}
$$

```{r}
sortino = port_daily[, (mean(ret) - rf) / sqrt(mean(pmin(ret - rf, 0)^2)) * sqrt(252)]
sortino
```

#### Rolling Sharpe

```{r}
#| dpi: 300
port_daily[, roll_sharpe := (frollmean(ret, window) - rf) / frollsd(ret, window) * sqrt(252)]

port_daily |>
  na.omit("roll_sharpe") |>
  ggplot(aes(x = date, y = roll_sharpe)) +
  geom_line() +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(title = "Rolling Sharpe Ratio (63-day)") +
  theme_finance
```

#### Value at Risk

Historical VaR at the 95% and 99% confidence levels:

```{r}
port_daily[, .(VaR_95 = quantile(ret, 0.05), VaR_99 = quantile(ret, 0.01))]
```

#### Expected Shortfall (CVaR)

Average loss beyond VaR:

```{r}
port_daily[, .(
  CVaR_95 = mean(ret[ret <= quantile(ret, 0.05)]),
  CVaR_99 = mean(ret[ret <= quantile(ret, 0.01)])
)]
```

#### Portfolio risk

Portfolio risk is defined as:

$$
\sigma_p = \sqrt{w^T \Sigma w}
$$

```{r}
cov_mat = dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cov(use = "pairwise.complete.obs")
wgt = alloc[colnames(cov_mat), weight, on = "ticker"]
port_risk = as.numeric(sqrt(t(wgt) %*% cov_mat %*% wgt))
port_risk
```

#### Drawdown

Maximum Drawdown is defined as follows:

$$
MDD = \max_{i \leq j} \left( \frac{V_i - V_j}{V_i} \right)
$$

Instrument drawdown:

```{r}
drawdown = copy(dt) |>
  _[, cum_ret := cumprod(1 + ret) - 1, by = ticker] |>
  _[, drawdown := (1 + cum_ret) / cummax(1 + cum_ret) - 1, by = ticker]
head(drawdown)
```

Portfolio drawdown:

```{r}
drawdown = dt |>
  _[, .(wret = sum(wret)), by = date] |>
  _[, cum_ret := cumprod(1 + wret) - 1] |>
  _[, drawdown := (1 + cum_ret) / cummax(1 + cum_ret) - 1]
head(drawdown)
```

```{r}
drawdown[drawdown < 0, .(min_drawdown = min(drawdown), avg_drawdown = mean(drawdown))]
```

#### Calmar ratio

Annualized return divided by maximum drawdown:

```{r}
calmar = port_daily[, (mean(ret) * 252) / abs(drawdown[, min(drawdown)])]
calmar
```

#### Tracking error

Tracking error measures how closely a portfolio follows its benchmark:

$$
TE = \sqrt{\frac{1}{N-1} \sum_{i=1}^{N} (r_{p,i} - r_{b,i})^2}
$$

```{r}
te = dt |>
  _[, .(port_ret = sum(wret)), by = date] |>
  _[bmr[, .(date, bmr_ret = ret)], on = "date", nomatch = NULL] |>
  _[, diff := port_ret - bmr_ret]

te[, .(
  daily_te = sd(diff),
  annual_te = sd(diff) * sqrt(252)
)]
```

#### Information ratio

Excess return per unit of tracking error:

```{r}
te[, mean(diff) / sd(diff) * sqrt(252)]
```

#### Beta and Alpha

Beta measures the portfolio's sensitivity to the benchmark, Alpha the excess return:

$$
r_p = \alpha + \beta \cdot r_b + \epsilon
$$

```{r}
fit = te[, lm(port_ret ~ bmr_ret)]
coefs = coef(fit)
data.table(alpha = coefs[1L] * 252, beta = coefs[2L])
```

#### Multi-factor model

The single-factor model above explains returns with the benchmark alone. Each instrument
is in fact driven by a common market factor plus a sector factor, so we can regress every
ticker on both and recover its factor loadings. Assemble the factors and reshape them to
long form, then match each ticker to its own sector factor with a join on `date` and
`sector`:

```{r}
factors = data.table(
  date = dates,
  mkt = market_ret,
  technology = sector_ret$technology,
  consumer_cyclical = sector_ret$consumer_cyclical
) |>
  melt(id.vars = c("date", "mkt"), variable.name = "sector", value.name = "sec_ret") |>
  _[, sector := as.character(sector)]

reg = factors[dt, on = .(date, sector), nomatch = NULL]
```

Run one regression per ticker and compare the recovered market beta with the value that
was used to generate the data:

```{r}
loadings = reg[, as.list(coef(lm(ret ~ mkt + sec_ret))), by = ticker] |>
  setnames(c("ticker", "alpha", "mkt_beta", "sector_beta")) |>
  _[, alpha := alpha * 252]
loadings[alloc, on = "ticker", true_beta := i.beta]
loadings
```

#### Rolling market beta

Beta is not static. Estimate it over a rolling window from the ratio of the rolling
covariance with the market to the market variance:

```{r}
#| dpi: 300
reg |>
  setorder(ticker, date) |>
  _[, let(
    cov_rm = frollmean(ret * mkt, window) - frollmean(ret, window) * frollmean(mkt, window),
    var_m = frollmean(mkt^2, window) - frollmean(mkt, window)^2
  ), by = ticker] |>
  _[, roll_beta := cov_rm / var_m] |>
  na.omit("roll_beta") |>
  ggplot(aes(x = date, y = roll_beta, color = ticker)) +
  geom_line() +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Rolling Market Beta (63-day)") +
  theme_finance +
  theme(legend.position = "bottom")
```

#### Correlation matrix

```{r}
cor_mat = dt |>
  dcast(date ~ ticker, value.var = "log_ret") |>
  _[, date := NULL] |>
  cor(use = "pairwise.complete.obs")
round(cor_mat, 3)
```

```{r}
#| dpi: 300
cor_dt = as.data.table(cor_mat, keep.rownames = "ticker1") |>
  melt(id.vars = "ticker1", variable.name = "ticker2", value.name = "cor")

ggplot(cor_dt, aes(x = ticker1, y = ticker2, fill = cor)) +
  geom_tile() +
  geom_text(aes(label = round(cor, 2)), size = 3) +
  scale_fill_gradient2(low = "#FF0000", mid = "white", high = "darkblue", midpoint = 0) +
  labs(title = "Correlation Matrix") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```
